---
title: "MedInfNetworks Networks"
output:
  html_document:
    toc: yes
    toc_float: yes
editor_options: 
  markdown: 
    wrap: 80
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

Please make sure you have the packages `CARNIVAL` (version 1.x.x and not 2.x.x) and `cosmosR` from Bioconductor installed. If you are using the Docker image, you will need to downgrade carnival to version 1.x.x.

```{r echo=TRUE, eval=FALSE}
BiocManager::install("cosmosR")
remotes::install_github("saezlab/CARNIVAL@v1.3")
```

This is a good time to go over the [previous practical](MedInfNetworks2021_Explore.Rmd) again to refresh your understanding of footprint based methods and the inputs that will be used for the network reconstruction methods that you will work with in this practical.


## 1) CARNIVAL

We proceed by contextualizing the estimations from the previous practicle in a causal network, combing prior knowledge networks and available data.

**Follow step 5**

<https://github.com/saezlab/transcriptutorial/blob/master/scripts/05_network_reconstruction_with_CARNIVAL.md>

To further analyse the results of running CARNIVAL

**Follow step 6**

<https://github.com/saezlab/transcriptutorial/blob/master/scripts/06_analysis_CARNIVAL_results.md>

If you are having troubles installing/downgrading `CARNIVAL` 1.x.x, with version 2.x.x. you can follow the `CARNIVAL` tutorial and run the toy examples 

<https://saezlab.github.io/CARNIVAL/articles/CARNIVAL.html>


```{r include=FALSE}
library(CARNIVAL)
load(file = system.file("toy_perturbations_ex1.RData",
                        package = "CARNIVAL"))
load(file = system.file("toy_measurements_ex1.RData",
                        package = "CARNIVAL"))
load(file = system.file("toy_network_ex1.RData",
                        package = "CARNIVAL"))

carnivalOptions <- defaultLpSolveCarnivalOptions()

# lpSolveresults
resultsLpSolve <- runVanillaCarnival( perturbations = toy_perturbations_ex1, 
                                      measurements = toy_measurements_ex1,
                                      priorKnowledgeNetwork = toy_network_ex1, 
                                      carnivalOptions = carnivalOptions)
```


Here is also a small script for visualizing the `CARNIVAL` outputs based on `visNetwork`.


```{r message=FALSE, warning=FALSE}
library(dplyr)
library(visNetwork)

resultsLpSolve

nodes <- data.frame(id = resultsLpSolve$nodesAttributes$Node, 
                    label = resultsLpSolve$nodesAttributes$Node, 
                    title = resultsLpSolve$nodesAttributes$nodesType)

edges <- resultsLpSolve$weightedSIF %>% 
  transmute(from = Node1, 
            to = Node2, 
            color = ifelse(Sign == 1, "blue", "red"), 
            value = Weight,
            arrows = "to")

visNetwork(nodes, edges)

```


## 2) COSMOS

COSMOS extends the anlaysis that can be performed with CARNIVAL and allows for integration of multiple modalities of measured data and not only transcriptomics.

Follow the cosmosR tutorial and run the toy example

<https://saezlab.github.io/cosmosR/articles/tutorial.html>

For more information see also the paper

```{r}
citation("cosmosR")
```

