---
title: "MedInfNetworks Explore"
output:
  html_document:
    toc: yes
    toc_float: yes
editor_options: 
  markdown: 
    wrap: 80
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

Follow the first part of
"[transcriptutorial](https://github.com/saezlab/transcriptutorial/)" to get
acquainted with the basics of data preprocessing, exploration and analysis with
footprint based methods.

## 1) Normalization and variance stabilization

Log transformation and variance stabilizing normalization with
[vsn](https://bioconductor.org/packages/vsn/).

See also the paper

```{r}
citation("vsn")
```

**Follow step 1**
<https://github.com/saezlab/transcriptutorial/blob/master/scripts/01_normalisation.md>

*Discussion: Other approaches to normalization and variance stabilization.*

## 2) Differential expression analysis

Differential expression analysis with
[limma](https://bioconductor.org/packages/limma/).

Consider skimming the
[userguide](https://bioconductor.org/packages/release/bioc/vignettes/limma/inst/doc/usersguide.pdf).
See also the paper

```{r}
citation("limma")
```

**Follow step 2**
<https://github.com/saezlab/transcriptutorial/blob/master/scripts/02_differential_analysis.md>

*Discussion: Assumptions of underlying distribution. Other approaches to
differential expression analysis.*

*(optional)* More on this topic [RNA-seq analysis is easy as 1-2-3 with limma,
Glimma and edge](https://f1000research.com/articles/5-1408)

## 3) Footprint based analyses

What is a footprint?

### PROGENy

See also the paper(s)

```{r}
citation("progeny")
```

**Follow step 3**
<https://github.com/saezlab/transcriptutorial/blob/master/scripts/03_Pathway_activity_with_Progeny.md>

### DoRothEA

See also the paper(s)

```{r}
citation("dorothea")
```

**Follow step 4**
<https://github.com/saezlab/transcriptutorial/blob/master/scripts/04_TranscriptionFactor_activity_with_Dorothea.md>

*Discussion: [decoupleR](https://saezlab.github.io/decoupleR/)*
